Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-12-19 manuscript_latex/manuscript.tex
Update manuscript for revision 3.

\begin{definition}[Numeric feature]
A numeric feature is a finite sequence $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in  \right\}$ of length $n$.
\end{definition}

@@

\begin{definition}[Central normality test]
The null-hypothesis $\mathcal{H}_0$ is that the sequence $\mathbf{X}$ was sampled
from a population that is distributed according to a central normal distribution.

The alternative hypothesis is that the sequence $\mathbf{X}$ was sampled from a
population that is not distributed according to a central normal distribution.
\end{definition}

@@

To assess whether the proposed power transformations lead to values of
\(\lambda\) that are invariant to location and scale of the
distribution, we simulated three different sequences. We first randomly
drew \(10000\) values from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed generalised normal
distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed generalised normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).
We then computed transformation parameter \(\lambda\) using the original
definitions (Eqn. \ref{eqn:box-cox-original} and
\ref{eqn:yeo-johnson-original}) and the location- and scale-invariant
definitions (Eqn. \ref{eqn:box-cox-invariant} and
\ref{eqn:yeo-johnson-invariant}) for each distribution. To assess
location invariance, a positive value \(d_{\text{shift}}\) was added to
each distribution with \(d_{\text{shift}} \in [1, 10^6]\). Similarly, to
assess scale invariance, each distribution was multiplied by a positive
value \(d_{\text{scale}}\), where \(d_{\text{scale}} \in [1, 10^6]\).

@@

The critical test statistic for the empirical central normality test
depends on the significance level \(\alpha\), sequence length \(n\) and
central portion \(\kappa\). Critical test statistic values were set
through simulation, as described below.

For each number of samples
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}\)
we randomly sampled \(m_d = 30000\) ordered sequences \(\mathbf{X}\)
from \(\mathcal{N}(0,1)\). For each sequence we then computed
\(\tau_{\kappa}\) for
\(\kappa \in \left\{0.50, 0.60, 0.70, 0.80, 0.90, 0.95, 1.00\right\}\).

@@

To assess type I error rates for the empirical central normality test
and compare these to the Shapiro-Wilk test for normality, we randomly
drew another 10000 sequences from a central normal distribution for each
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}\).

Each sequence was created as follows: first \(n\) values are sampled
from \(\mathcal{N}(0, 1)\). Then, the elements outside the central
portion of the sequence
(i.e.~\(\mathbf{X} \setminus \mathbf{X}_{\text{c}}\)) are replaced.
Elements with \(p_i < \frac{1-\kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))\), and
elements with \(p_i > \frac{1 + \kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)\).

For each sequence, the p-value for the respective test was used to
reject the null hypothesis that sequences were derived from a population
with a (central) normal distribution. The null hypothesis was rejected
if \(p \leq 0.05\). The type I error rate was determined by computing
the fraction of sequences rejected this way. Figure
\ref{fig:empirical-central-normality-error_rate} shows that when the
central portion \(\kappa\) of the central normal distribution is equal
or greater than the \(\kappa\) parameter for the empirical central
normality test, the tests are accurate. If \(\kappa\) of the empirical
central normality test exceeds that the central portion \(\kappa\) for
the underlying distribution, tests are conservative. The Shapiro-Wilk
test behaves similar to the empirical central normality test with
\(\kappa = 1.0\), which is expected. For \(n < 10\), the empirical
central normality test can be optimistic.

@@

To determine the weighting function parameters for each of the nine
combinations, \(m_d=500\) sequences \(\mathbf{X}_i\)
(\(i \in \{1, 2, \ldots, m_d\}\)) were randomly drawn from randomly
parametrised asymmetric generalised normal distributions. Each
distribution was parametrised with a random skewness parameter
\(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
\(\beta \sim U\left(1.00, 5.00 \right)\). Location and scale parameters
were set as \(\mu = 0\) and \(\sigma = 1\), respectively. To form each
sequence \(\mathbf{X}_i\), \(n = \lceil 10^\gamma \rceil\) instances
were then randomly drawn, with
\(\gamma \sim U\left(\log_{10}50, 3\right)\), resulting in a set of
sequences with between \(50\) and \(1000\) elements each. Outlier values
were then drawn to randomly replace 10 percent of the elements of
\(\mathbf{X}_i\). These values were set according to
\citet{Tukey1977-xm}, as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

@@

To find the optimal values for the weighting function parameters
\(\delta_1\) and \(\delta_2\) (if applicable), we minimised a composite
loss
\(L = \sum_{i=1}^{m_d} L_{\text{cn},i} + 0.1 \sum_{i=1}^{m_d} L_{\lambda,i}\).
The composite loss consisted of two components: a loss term
\(L_{\text{cn},i} = \tau_{n_i, \kappa = 0.80,i}\), i.e.~the mean of
residual errors of the central 80\% of elements of each sequence
\(\mathbf{X}_i\), which aimed at optimising central normality; and a
loss term
\(L_{\lambda,i} = \max\left(0.0,|\lambda_{0,i} - \lambda_{i}| - \xi \right)\),
with \(\lambda_{0,i}\) and \(\lambda_{i}\) transformation parameters
found for sequence \(\mathbf{X}_i\) prior to and after adding outliers,
respectively, and tolerance parameter \(\xi = 0.5\) for Box-Cox and
\(\xi = 0.3\) for Yeo-Johnson power transformations. This second term
aimed to prevent solutions that provide small improvements in central
normality at the cost of a poor fit of the tails of sequences.

@@

Three datasets with 100000 sequences each were created to assess
transformation to normality. For each sequence \(\mathbf{X}_i\),
\(n = \lceil 10^\gamma \rceil\) elements were randomly drawn, with
\(\gamma \sim U\left(1, 4\right)\), resulting in sequences with between
\(10\) and \(10000\) elements.

\begin{itemize}
\item
  A \textit{clean} dataset with each sequence drawn from
  \(\mathcal{N}(0, 1)\) and transformed using an inverse power
  transformation with randomly drawn transformation parameter
  \(\lambda \sim U\left(0.00, 2.00 \right)\).
  \(\left(\phi_{\text{BC}}^\lambda\right)^{-1}\) and
  \(\left(\phi_{\text{YJ}}^\lambda\right)^{-1}\) were used as inverse
  transformations for assessing Box-Cox and Yeo-Johnson transformations,
  respectively. Prior to inverse transformation, sequences for Box-Cox
  transformations were shifted into the positive domain, with minimum
  value \(1\).
\item
  A \textit{dirty} dataset with each sequence drawn from
  \(\mathcal{AGN}\left(\mu = 0, \sigma = 1/\sqrt{2}, \alpha, \beta \right)\),
  with randomly drawn skewness parameter
  \(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
  \(\beta \sim U\left(1.00, 5.00 \right)\).
\item
  A \textit{shifted} dataset with each sequence drawn from
  \(\mathcal{AGN}\left(\mu = 100, \sigma = 10^{-3} \cdot 1/\sqrt{2} , \alpha, \beta \right)\),
  with randomly drawn skewness parameter
  \(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
  \(\beta \sim U\left(1.00, 5.00 \right)\).
\end{itemize}

A dataset with outliers was created for each of the above datasets by
replacing 10 percent of elements in each sequence, as described earlier.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of sequence
  \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / Q_n\left(\mathbf{X}\right)\),
  with \(Q_n\) representing the robust \(Q_n\) scale estimator of
  \(\mathbf{X}\) \citep{Croux1992-zz, Rousseeuw1993-it}.
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 100000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulations_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

\begin{itemize}
\item
  very easy: \(\text{AUC} \geq 0.90\) or \(\text{RRSE} \leq 0.10\) (57
  datasets)
\item
  easy: \(0.90 > \text{AUC} \geq 0.80\) or
  \(0.30 \geq \text{RRSE} > 0.10\) (40 datasets)
\item
  intermediate: \(0.80 > \text{AUC} \geq 0.70\) or
  \(0.60 \geq \text{RRSE} > 0.30\) (52 datasets)
\item
  difficult: \(0.70 > \text{AUC} \geq 0.60\) or
  \(0.80 \geq \text{RRSE} > 0.60\) (33 datasets)
\item
  very difficult: \(0.60 > \text{AUC} \geq 0.50\) or
  \(1.00 \geq \text{RRSE} > 0.80\) (48 datasets)
\item
  unsolvable: \(\text{AUC} < 0.50\) or \(\text{RRSE} > 1.00\) (1
  dataset)
\end{itemize}

@@

We introduced an empirical test for central normality to assess whether
sequences deviate from normality in a way that might require closer
inspection prior to further processing. The empirical central normality
test differs from other tests for normality, such as the Shapiro-Wilk
test \citep{Shapiro1965-zd}, as it assesses normality of the central
portion of a feature, instead of the entire feature. Compared to the
Shapiro-Wilk test, the empirical central normality test remains
consistent when a feature was not sampled from a normal distribution,
such as a central normal distribution with \(\kappa < 1.0\). This
enables assessing reasonable normality of (transformed) features in the
presence of outliers.

This work has the following limitation: We observed several numerical
stability issues for optimisation criteria other than MLE (\hyperref[appendix-b]{Appendix B}).
These appear in regions where transformation parameters would lead to
very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed.

@@

\begin{table}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
If necessary, each sequence was shifted so that every element had a strictly positive value.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

In the main manuscript, transformation methods are compared by ranking
within each experiment, defined by the combination of sequence and
absence/presence of outliers. This allows for direct comparison of
methods between different experiments.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
shows the average normalised residual error for these transformation
methods. This metric is computed from the residual error and the
residual error of the central portion (\(\kappa = 0.80\)) for sequences
without and with outliers, respectively. In each experiment, the
respective residual error was normalised by dividing with the sequence
length, and can be interpreted as the average residual error of a single
sample. This reduces the effect of sequence length on the residual
error.

Table
\ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
shows the average normalised difference in residual error between the
best method for each experiment and the individual methods. As above,
normalisation was done by division by the length of the sequence
corresponding to each experiment.

@@

\begin{table}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\caption{
Comparison of average difference in residual error between Yeo-Johnson
transformation methods and the best transformation method within each
experiment, normalised by the corresponding sequence length. This is based on
either the residual error (without outliers) or the residual error of the
central portion (with outliers; $\kappa = 0.80$) over 3 datasets with 100000
sequences each. Residual errors were normalised by sequence length to reduce the
effect of sequence length on the residual error. The clean dataset consists of
sequences derived through inverse Yeo-Johnson transformation of data sampled
from a standard normal distribution. The dirty dataset contains sequences
sampled from asymmetric generalised normal distributions, centred at 0. The
shifted dataset also contains sequences sampled from asymmetric generalised
normal distributions, but centred at 100, and scaled by 0.001. Several
transformation methods include normalisation before transformation, indicated by
z-score normalisation (norm.) or robust scaling. A lower difference is better,
with a minimum value of 0.0. Discrepancies with the rank-based method may occur,
as normalised differences in residual error are not fully comparable between
experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
\begin{tabular}{l | l r r r r r r}

@@

\subsection{Ranking Yeo-Johnson transformations: effect of sequence
length}\label{ranking-yeo-johnson-transformations-effect-of-sequence-length}

>>>>

\begin{definition}[Numeric feature]
A numeric feature is a vector $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R} \right\}$ of length $n$.
\end{definition}

@@

If $\mathbf{X}$ is randomly sampled from a normal distribution,
$\sup_{r_j \in \mathbf{R}_{\text{c}}} r_j \rightarrow 0$, almost surely.
That is, every residual of the central portion of $\mathbf{X}$ will asymptotically
converge to 0.

The latter corollary follows from the definition of the residuals and
the restriction to normal distributions. Residuals were defined using
\(\frac{x_i - \mu_M}{\sigma_M} \approx \hat{F}^{-1}_{\mathcal{N}}(p_i)\),
with \(\hat{F}^{-1}_{\mathcal{N}}(p_i)\) the empirical quantile function
resembling the quantile function of the normal distribution
\(\mathcal{N}(0, 1)\). Since
\(\hat{F}^{-1}_{\mathcal{N}}(p) \rightarrow F^{-1}_{\mathcal{N}}(p)\)
with \(\frac{1-\kappa}{2} \leq  p \leq \frac{1 + \kappa}{2}\)
\citep{van-der-Vaart1998-xu},
\(\sup_{r_j \in \mathbf{R}_{\text{c}}} r_j \rightarrow 0\), if and only
if for all \(x_i \in \mathbf{X}_c\),
\(\frac{x_i - \mu_M}{\sigma_M} \rightarrow \hat{F}^{-1}_{\mathcal{N}}(p_i)\).
Thus, the requirement is that the estimators for location and scale
\(\mu_M\) and \(\sigma_M\) are bias-free. M-estimators for location and
scale satisfy this condition, if \(\mathbf{X}\) follows a normal
distribution without contamination \citep{Huber1964-xk, Huber1981-su}.

\begin{definition}[Central normality test]
The null-hypothesis $\mathcal{H}_0$ is that $\mathbf{X}$ was randomly sampled
from a population that is distributed according to a central normal distribution.

The alternative hypothesis is that $\mathbf{X}$ was randomly sampled from a
population that is not distributed according to a central normal distribution.
\end{definition}

@@

To assess whether the proposed power transformations lead to values of
\(\lambda\) that are invariant to location and scale of the
distribution, we simulated three different numeric features. We first
randomly drew \(10000\) values from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed generalised normal
distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed generalised normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).
We then computed transformation parameter \(\lambda\) using the original
definitions (Eqn. \ref{eqn:box-cox-original} and
\ref{eqn:yeo-johnson-original}) and the location- and scale-invariant
definitions (Eqn. \ref{eqn:box-cox-invariant} and
\ref{eqn:yeo-johnson-invariant}) for each distribution. To assess
location invariance, a positive value \(d_{\text{shift}}\) was added to
each distribution with \(d_{\text{shift}} \in [1, 10^6]\). Similarly, to
assess scale invariance, each distribution was multiplied by a positive
value \(d_{\text{scale}}\), where \(d_{\text{scale}} \in [1, 10^6]\).

@@

The critical test statistic for the empirical central normality test
depends on the significance level \(\alpha\), feature length \(n\) and
central portion \(\kappa\). Critical test statistic values were set
through simulation, as described below.

For each number of samples
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}\)
we randomly sampled \(m_d = 30000\) ordered numeric features
\(\mathbf{X}\) from \(\mathcal{N}(0,1)\). For each feature we then
computed \(\tau_{\kappa}\) for
\(\kappa \in \left\{0.50, 0.60, 0.70, 0.80, 0.90, 0.95, 1.00\right\}\).

@@

To assess type I error rates for the empirical central normality test
and compare these to the Shapiro-Wilk test for normality, we randomly
drew another 10000 numeric features from a central normal distribution
for each
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}\).

Each feature was created as follows: first \(n\) values are sampled from
\(\mathcal{N}(0, 1)\). Then, the elements outside the central portion of
the feature (i.e.~\(\mathbf{X} \setminus \mathbf{X}_{\text{c}}\)) are
replaced. Elements with \(p_i < \frac{1-\kappa}{2}\) are replaced with
values sampled from \(\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))\),
and elements with \(p_i > \frac{1 + \kappa}{2}\) are replaced with
values sampled from \(\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)\).

For each feature, the p-value for the respective test was used to reject
the null hypothesis that features were derived from a population with a
(central) normal distribution. The null hypothesis was rejected if
\(p \leq 0.05\). The type I error rate was determined by computing the
fraction of features rejected this way. Figure
\ref{fig:empirical-central-normality-error_rate} shows that when the
central portion \(\kappa\) of the central normal distribution is equal
or greater than the \(\kappa\) parameter for the empirical central
normality test, the tests are accurate. If \(\kappa\) of the empirical
central normality test exceeds that the central portion \(\kappa\) for
the underlying distribution, tests are conservative. The Shapiro-Wilk
test behaves similar to the empirical central normality test with
\(\kappa = 1.0\), which is expected. For \(n < 10\), the empirical
central normality test can be optimistic.

@@

To determine the weighting function parameters for each of the nine
combinations, \(m_d=500\) numeric features \(\mathbf{X}_i\)
(\(i \in \{1, 2, \ldots, m_d\}\)) were randomly drawn from randomly
parametrised asymmetric generalised normal distributions. Each
distribution was parametrised with a random skewness parameter
\(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
\(\beta \sim U\left(1.00, 5.00 \right)\). Location and scale parameters
were set as \(\mu = 0\) and \(\sigma = 1\), respectively. To form each
feature \(\mathbf{X}_i\), \(n = \lceil 10^\gamma \rceil\) instances were
then randomly drawn, with \(\gamma \sim U\left(\log_{10}50, 3\right)\),
resulting in a set of features with between \(50\) and \(1000\) elements
each. Outlier values were then drawn to randomly replace 10 percent of
the elements of \(\mathbf{X}_i\). These values were set according to
\citet{Tukey1977-xm}, as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

@@

To find the optimal values for the weighting function parameters
\(\delta_1\) and \(\delta_2\) (if applicable), we minimised a composite
loss
\(L = \sum_{i=1}^{m_d} L_{\text{cn},i} + 0.1 \sum_{i=1}^{m_d} L_{\lambda,i}\).
The composite loss consisted of two components: a loss term
\(L_{\text{cn},i} = \tau_{n_i, \kappa = 0.80,i}\), i.e.~the mean of
residual errors of the central 80\% of elements of each feature
\(\mathbf{X}_i\), which aimed at optimising central normality; and a
loss term
\(L_{\lambda,i} = \max\left(0.0,|\lambda_{0,i} - \lambda_{i}| - \xi \right)\),
with \(\lambda_{0,i}\) and \(\lambda_{i}\) transformation parameters
found for feature \(\mathbf{X}_i\) prior to and after adding outliers,
respectively, and tolerance parameter \(\xi = 0.5\) for Box-Cox and
\(\xi = 0.3\) for Yeo-Johnson power transformations. This second term
aimed to prevent solutions that provide small improvements in central
normality at the cost of a poor fit of the tails of features.

@@

Three datasets with 100000 numeric features each were created to assess
transformation to normality. For each feature \(\mathbf{X}_i\),
\(n = \lceil 10^\gamma \rceil\) elements were randomly drawn, with
\(\gamma \sim U\left(1, 4\right)\), resulting in features with between
\(10\) and \(10000\) elements.

\begin{itemize}
\item
  A \textit{clean} dataset with each feature drawn from
  \(\mathcal{N}(0, 1)\) and transformed using an inverse power
  transformation with randomly drawn transformation parameter
  \(\lambda \sim U\left(0.00, 2.00 \right)\).
  \(\left(\phi_{\text{BC}}^\lambda\right)^{-1}\) and
  \(\left(\phi_{\text{YJ}}^\lambda\right)^{-1}\) were used as inverse
  transformations for assessing Box-Cox and Yeo-Johnson transformations,
  respectively. Prior to inverse transformation, features for Box-Cox
  transformations were shifted into the positive domain, with minimum
  value \(1\).
\item
  A \textit{dirty} dataset with each feature drawn from
  \(\mathcal{AGN}\left(\mu = 0, \sigma = 1/\sqrt{2}, \alpha, \beta \right)\),
  with randomly drawn skewness parameter
  \(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
  \(\beta \sim U\left(1.00, 5.00 \right)\). The features are drawn from
  a range of asymmetric unimodal distributions to reflect a wider range
  of generating distributions than for the data presented in the first
  dataset.
\item
  A \textit{shifted} dataset with each feature drawn from
  \(\mathcal{AGN}\left(\mu = 1000, \sigma = 1/\sqrt{2} , \alpha, \beta \right)\),
  with randomly drawn skewness parameter
  \(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
  \(\beta \sim U\left(1.00, 5.00 \right)\). The generating process
  creates features similar to those in the \textit{dirty} dataset, but
  shifted to a location that leads to normalisation issues observed for
  conventional power transformation methods, as shown in Figure
  \ref{fig:decreased-normality}.
\end{itemize}

A dataset with outliers was created for each of the above datasets by
replacing 10 percent of elements in each feature, as described earlier.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / Q_n\left(\mathbf{X}\right)\),
  with \(Q_n\) representing the robust \(Q_n\) scale estimator of
  \(\mathbf{X}\) \citep{Croux1992-zz, Rousseeuw1993-it}.
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each feature. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each feature after transformation.
Then, each method was ranked according to the sum of residual errors
(data without outliers) or the sum of residual errors of the central
portion (data with outliers). The average rank of each transformation
method was computed over all 100000 features in each dataset. Average
ranks for Yeo-Johnson transformations are shown in Table
\ref{tab:comparison_methods_simulations_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 features each. The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

\begin{itemize}
\tightlist
\item
  very easy: \(\text{AUC} \geq 0.90\) or \(\text{RRSE} \leq 0.10\) (57
  datasets)
\item
  easy: \(0.90 > \text{AUC} \geq 0.80\) or
  \(0.30 \geq \text{RRSE} > 0.10\) (40 datasets)
\item
  intermediate: \(0.80 > \text{AUC} \geq 0.70\) or
  \(0.60 \geq \text{RRSE} > 0.30\) (52 datasets)
\item
  difficult: \(0.70 > \text{AUC} \geq 0.60\) or
  \(0.80 \geq \text{RRSE} > 0.60\) (33 datasets)
\item
  very difficult: \(0.60 > \text{AUC} \geq 0.50\) or
  \(1.00 \geq \text{RRSE} > 0.80\) (48 datasets)
\item
  unsolvable: \(\text{AUC} < 0.50\) or \(\text{RRSE} > 1.00\) (1
  dataset)
\end{itemize}

@@

We introduced an empirical test for central normality to assess whether
features deviate from normality in a way that might require closer
inspection prior to further processing. The empirical central normality
test differs from other tests for normality, such as the Shapiro-Wilk
test \citep{Shapiro1965-zd}, as it assesses normality of the central
portion of a feature, instead of the entire feature. Compared to the
Shapiro-Wilk test, the empirical central normality test remains
consistent when a feature was not sampled from a normal distribution,
such as a central normal distribution with \(\kappa < 1.0\). This
enables assessing reasonable normality of (transformed) features in the
presence of outliers.

This work has the following limitations: We observed several numerical
stability issues for optimisation criteria other than MLE (\hyperref[appendix-b]{Appendix B}).
These appear in regions where transformation parameters would lead to
very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed. Moreover,
the definition of central normal distribution does not constrain the
distribution of the non-central tails. Thus, the test may be
conservative for central normal distributions other than uncontaminated
normal distributions, because the location and scale estimators are not
generally bias-free \citep{Rousseeuw1994-ur}.

@@

\begin{table}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 features each. The clean dataset consists of features derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
If necessary, each feature was shifted so that every element had a strictly positive value.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

In the main manuscript, transformation methods are compared by ranking
within each experiment, defined by the combination of feature and
absence/presence of outliers. This allows for direct comparison of
methods between different experiments.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
shows the average normalised residual error for these transformation
methods. This metric is computed from the residual error and the
residual error of the central portion (\(\kappa = 0.80\)) for features
without and with outliers, respectively. In each experiment, the
respective residual error was normalised by dividing with the feature
length, and can be interpreted as the average residual error of a single
sample. This reduces the effect of feature length on the residual error.

Table
\ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
shows the average normalised difference in residual error between the
best method for each experiment and the individual methods. As above,
normalisation was done by division by the feature length of each experiment.

@@

\begin{table}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 features each. Residual errors were normalised by feature length to reduce the effect of feature length on the residual error.
The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\caption{
Comparison of average difference in residual error between Yeo-Johnson
transformation methods and the best transformation method within each
experiment, normalised by the corresponding feature length. This is based on
either the residual error (without outliers) or the residual error of the
central portion (with outliers; $\kappa = 0.80$) over 3 datasets with 100000
features each. Residual errors were normalised by feature length to reduce the
effect of feature length on the residual error. The clean dataset consists of
features derived through inverse Yeo-Johnson transformation of data sampled
from a standard normal distribution. The dirty dataset contains features
sampled from asymmetric generalised normal distributions, centred at 0. The
shifted dataset also contains features sampled from asymmetric generalised
normal distributions, but centred at 1000. Several
transformation methods include normalisation before transformation, indicated by
z-score normalisation (norm.) or robust scaling. A lower difference is better,
with a minimum value of 0.0. Discrepancies with the rank-based method may occur,
as normalised differences in residual error are not fully comparable between
experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
\begin{tabular}{l | l r r r r r r}

@@

\subsection{Ranking Yeo-Johnson transformations: effect of feature
length}\label{ranking-yeo-johnson-transformations-effect-of-feature-length}
